---
title: "English Team of the Year Award Analysis"
output: html_notebook
---

```{r include=FALSE}
library(tidyverse)
library(rvest)
library(stringr)
library(scales)
library(lubridate)
library(ggthemes)
library(reshape2)
```


Step 1:
Web Scraping of 1970s, 1980s, 1990s, 2000s, and 2010s Wikipedia pages for Team of the Year

```{r full decades 1980s-2000s, include=FALSE}

#initialize data.frame
df <- data.frame(`Pos.`= as.character(),
                 Player = as.character(),
                 Club = as.character(),
                 `App.` = as.double(),
                 year = as.character())


div_table_numbers <- c(1,5,9,13,17,21,25,29,33,37)

urls <- c("https://en.wikipedia.org/wiki/PFA_Team_of_the_Year_(2000s)",
  "https://en.wikipedia.org/wiki/PFA_Team_of_the_Year_(1990s)",
  "https://en.wikipedia.org/wiki/PFA_Team_of_the_Year_(1980s)")

for (i in 1:length(urls)){
  
  for (j in 1:length(div_table_numbers)) {
  
    xpath_base <- '//*[@id="mw-content-text"]/div/table['
    new_data <- urls[i] %>%
      html() %>%
      html_nodes(xpath = paste0(xpath_base,
                              div_table_numbers[j],
                              "]")) %>%
      html_table()
  
    new_data <- new_data[[1]]
    #new_data
    
    #get year
    year_xpath_base <- '//*[@id="mw-content-text"]/div/h3['
    year <- urls[i] %>%
      html() %>%
      html_nodes(xpath = paste0(year_xpath_base,
                              j,
                              "]")) %>%
      html_text()
    
    #remove "[edit]" text which Wikipedia includes
    year <- year %>% str_remove_all(fixed("[edit]"))
    year
    
    
    new_data$year <- year
    
    df <- bind_rows(df,
                    new_data)
  
  }
}

```

```{r partial decadec1970s, include=FALSE}

url_1970s <- "https://en.wikipedia.org/wiki/PFA_Team_of_the_Year_(1970s)"

seventies_div_table_numbers <- c(1,5,9,13,17,21)

for (j in 1:length(seventies_div_table_numbers)) {
  
    xpath_base <- '//*[@id="mw-content-text"]/div/table['
    new_data <- url_1970s %>%
      html() %>%
      html_nodes(xpath = paste0(xpath_base,
                              seventies_div_table_numbers[j],
                              "]")) %>%
      html_table()
  
    new_data <- new_data[[1]]
    
    #get year
    year_xpath_base <- '//*[@id="mw-content-text"]/div/h3['
    
    year <- url_1970s %>%
      html() %>%
      html_nodes(xpath = paste0(year_xpath_base,
                              j,
                              "]") )%>%
      html_text()
    
    #remove "[edit]" text which Wikipedia includes
    year <- year %>% str_remove_all(fixed("[edit]"))
    year
    
    
    new_data$year <- year
    df <- bind_rows(df,
                    new_data)
  
  }



```


```{r partial decade 2010s, include=FALSE}

url_2010s <- "https://en.wikipedia.org/wiki/PFA_Team_of_the_Year_(2010s)"

twenty_tens_div_table_numbers <- c(1,5,9,13,17,22,27,32,37)

for (j in 1:length(twenty_tens_div_table_numbers)) {
  
    xpath_base <- '//*[@id="mw-content-text"]/div/table['
    new_data <- url_2010s %>%
      html() %>%
      html_nodes(xpath = paste0(xpath_base,
                              twenty_tens_div_table_numbers[j],
                              "]")) %>%
      html_table()
  
    #create new_data df from new_data list
    new_data <- new_data[[1]]
    
    #get year
    year_xpath_base <- '//*[@id="mw-content-text"]/div/h3['
    year <- url_2010s %>%
      html() %>%
      html_nodes(xpath = paste0(year_xpath_base,
                              j,
                              "]"))%>%
      html_text()
    
    year <- year %>% str_remove_all(fixed("[edit]"))
    year
    
    
    new_data$year <- year
    df <- bind_rows(df,
                    new_data)
  
  }



```


Step 2: Exploratory Analysis

```{r seeing how formation has changed}

df <- df %>%
  mutate(short_year = str_sub(year,1,4) %>% as.numeric() + 1)

order_positions <- c("GK","DF","MF","FW")

df <- df %>% mutate(Pos. = fct_relevel(Pos., order_positions))



count_position <- df %>%
  filter(Pos. %in% c("MF","FW")) %>%
    count(Pos.,short_year, sort = TRUE)

count_position %>%
  ggplot(aes(x=short_year,y=n,color=Pos.,group=Pos.)) + 
  geom_point(position=position_jitter(h=0.005))+
  geom_smooth(method = "loess")+
  scale_x_continuous(breaks = seq(1973, 2017, 5))+
  scale_y_continuous(breaks=seq(2,4,1))+
  labs(x = "Year",
       y = "Number of Players per Position",
       title = "Count of Midfielders and Forwards in English PFA Team of the Year (1973 - 2017)",
       caption = "Data Source: Wikipedia")+
  theme_few()+
  theme(plot.title = element_text(hjust = 0.5))

ggsave(filename = "static_plot_players_per_formation.png",width = 30, height = 15, units = "cm")
#ggsave()
```

```{r}

club_count <- df %>% count(Club, sort = TRUE)

club_count %>% 
  top_n(n = 15) %>% 
  ggplot(aes(x = reorder(Club,n), 
                  y = n,
                  fill = Club)) +
  geom_col() +
  coord_flip()+
  theme_few()+
  guides(fill=FALSE)+
  labs(y = "# of Players",
       x = "Club",
       title = "Number of Players in PFA Team of the Year (1973 - 2017)",
       caption = "Data Source: Wikipedia")+
  theme(plot.title = element_text(hjust = 0.5))

#ggsave(filename = "static_plot_players_per_club.png",width = 30, height = 15, units = "cm")

```

```{r}

club_count_year <- df %>% 
  count(Club, year, sort = TRUE) %>%
  mutate(club_year = paste(Club, year))

club_count_year %>% 
  top_n(n = 15) %>% 
  ggplot(aes(x = reorder(club_year,n), 
                  y = n,
                  fill = club_year)) +
  geom_col() +
  coord_flip()+
  theme_few()+
  guides(fill=FALSE)+
  labs(y = "# of Players",
       x = "Club",
       title = "Number of Players in PFA Team of the Year (1973 - 2017)",
       caption = "Data Source: Wikipedia")+
  theme(plot.title = element_text(hjust = 0.5))





```

```{r}

by_position_top_performers <- df %>% count(Player,Pos.)

by_position_top_performers %>% 
  group_by(Pos.) %>% 
  top_n(n = 5, wt = n) %>%
  ggplot(aes(x = reorder(Player,n),
             y = n,
             fill = Pos.)) +
  geom_col() +
  facet_wrap(~Pos., scales = "free_y") +
  coord_flip() +
  theme_few() +
  scale_y_continuous(breaks=seq(0,10,2)) +
  guides(fill=FALSE) +
  labs(x = NULL,
       y = NULL,
       title = "Number of Appearances in PFA Team of the Year per Position (1973 - 2017)",
       caption = "\nData Source: Wikipedia")+
  theme(plot.title = element_text(hjust = 0.5))
  

ggsave(filename = "static_plot_players_appearances_number.png",width = 30, height = 15, units = "cm")

```



Winners

```{r}

#pre-EPL, First Division

first_div_top_three_url <- "https://en.wikipedia.org/wiki/List_of_English_football_champions"
first_div_top_three_xpath <- '//*[@id="mw-content-text"]/div/table[2]'

first_div_top_three <- first_div_top_three_url %>%
  html() %>%
  html_nodes(xpath = first_div_top_three_xpath) %>%
  html_table()
  
             
first_div_top_three <- first_div_top_three[[1]]           

first_div_top_three <- first_div_top_three %>% filter(!(Year %in% c("1915/16–1918/19",
                                                                    "1939/40–1945/46")))
first_div_top_three$Goals <-first_div_top_three$Goals %>% as.numeric()

first_div_top_three <- first_div_top_three %>% 
  rename(`Champions` = `Champions(number of titles)`,
         `Top goalscorer` = `Leading goalscorer`)

###EPL

epl_top_three_url <- "https://en.wikipedia.org/wiki/List_of_English_football_champions"
epl_top_three_xpath <- '//*[@id="mw-content-text"]/div/table[3]'

epl_top_three <- epl_top_three_url %>%
  html() %>%
  html_nodes(xpath = epl_top_three_xpath) %>%
  html_table()
  
             
epl_top_three <- epl_top_three[[1]] 

epl_top_three <- epl_top_three %>% 
  rename(`Champions` = `Champions (number of titles)`)

english_top_three_total <- bind_rows(first_div_top_three,
                                     epl_top_three)


english_top_three_total$Champions <- english_top_three_total$Champions %>% str_remove_all(regex("\\([^)]*\\)"))

english_top_three_total$Champions <- english_top_three_total$Champions %>% str_remove_all(regex("\\[.*?\\]"))

english_top_three_total <- english_top_three_total %>%
  mutate(short_year = str_sub(Year,1,4) %>% as.numeric() + 1)

english_top_three_total <- english_top_three_total %>%
  filter(short_year > 1973) %>%
  select(-c(`Top goalscorer`,Goals))

#melt
english_top_three_total_melted <- english_top_three_total %>%
  melt(id.vars=c("Year","short_year"),
       value.name = "Club",
       variable.name = "Team_Ranking")

english_top_three_total_melted$Club <- english_top_three_total_melted$Club %>% str_trim(side = c( "right"))

```

```{r merge datasets}

df_merged <- df %>% left_join(english_top_three_total_melted,
                              by = c("Club","short_year"))

```



* Teams That won the Premier League and Had Fewest Players
* Teams that didn't win the Premier League but had the most players included


How many players from champions were picked
```{r}

players_champions_per_year <- df_merged %>%
  filter(!is.na(Team_Ranking)) %>%
  count(short_year, Team_Ranking)
  
players_champions_per_year %>% ggplot(aes(x = short_year,
                                          y = n,
                                          color = Team_Ranking))+ geom_point(position=position_jitter(h=0.005))+
   scale_x_continuous(breaks = seq(1973, 2017, 2)) +
  theme_few()

```


Champions of England with the most player of the year awards
```{r}

champions_players <- df_merged %>%
  filter(Team_Ranking == "Champions") %>%
  count(Club, short_year, year, sort = TRUE) %>%
  mutate(club_year = paste(Club, year))

champions_players %>% 
  top_n(n = 10, wt = n) %>% 
  ggplot(aes(x = reorder(club_year,n), 
                  y = n,
                  fill = Club)) +
  geom_col() +
  coord_flip()+
  theme_few()+
  guides(fill=FALSE)+
  labs(y = "# of Players",
       x = "Club",
       title = "Champions of England w/ Most Players in PFA Team of the Year (1973 - 2017)",
       caption = "Data Source: Wikipedia")+
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values = c("Manchester United" = "darkred", 
                                 "Liverpool" = "orange",
                                 "Arsenal" = "yellow",
                                 "Chelsea" = "blue",
                                 "Blackburn Rovers" = "lightblue",
                                 "Leeds United" = "grey",
                                 "Manchester City" = "dark green"))


```


Winners with the least 
```{r}

```


Second or third place teams with the most 


Future work
